Palomar 5 Gaps

### 2024-11-07 paper/main.tex
moved to the end
added pearson and amorisco citations

\abstract
  {Holy moly artichokey}
 % aims heading (mandatory)
  {ok}
 % methods heading (mandatory)
  {good}
 % results heading (mandatory)
  {alright}
 % conclusions heading (optional), leave it empty if necessary
  {done}

\maketitle
\section{Introduction}
  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC 300 and M~33 cannot be fully explained by disk models and that there must be the same amount of mass or more that is undetected and beyond the extend of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat outside its effective radius and not decreasing as one would expect from a potential with more mass concentration in the center.

@@

  \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes theoretically and how to model them pratically, showcasing how different assumptions for the particle behavior of dark matter influences the mass and size distribution of sub haloes. The authors rightly state that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive the possibilities for sub-galactic behaviour and structure of dark matter are vaste. Obtaining constraints on the distirbution of dark matter subhaloes in turn will place constraints on the particle nature of dark matter. As proposed by \citet{2012ApJ...748...20C}, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. They are one of the evidences for the hierarchical nature of the galaxy formation as predicted by the $\Lambda CDM$ model \citep{1991ApJ...379...52W}. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of \citet{2024arXiv240519410B} properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

@@

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}.

  Previous studies have predominantly focused on the creation of gaps in stellar streams due to dark matter subhalo passages, with some exceptions that considered giant molecular clouds. CITE(AMORISCO 2016) CITE(Pearson et al 2017 BAR ON PAL 5) To our knowledge, no study has yet been performed investigating the gravitational impact of globular clusters on stellar streams. Since clusters are dense stellar systems, distributed mainly in the Galactic halo, they too should in principle contribute to generating perturbation on stellar streams. If globular clusters can indeed create inhomogeneities and gaps in streams, this factor must be taken into account when using stellar streams to infer the population of dark matter subhaloes.

>>>>

\abstract
  {ok}
 % aims heading (mandatory)
  {ok}
 % methods heading (mandatory)
  {ok}
 % results heading (mandatory)
  {ok}
 % conclusions heading (optional), leave it empty if necessary
  {done}

\maketitle
\section{Introduction}
  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC~300 and M~33 cannot be fully explained by disk models and that there must be the same amount of undetected mass as luminous mass and beyond the extent of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat outside its effective radius---not decreasing as one would expect from a gravitational field from a centrally concentrated mass distribution.

@@

  \citet{2019Galax...7...81Z} provide a review for the different proposed mechanisms of forming dark matter haloes theoretically and how to model them pratically, showcasing how different assumptions for the fundamental particle behavior of dark matter propogates to the mass and size distribution of sub haloes on the galactic scale. The authors rightly state that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive the possibilities for sub-galactic behaviour and structure of dark matter are vaste. Obtaining constraints on the distirbution of dark matter subhaloes in turn will place constraints on the particle nature of dark matter. As proposed by \citet{2012ApJ...748...20C}, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. They are one of the many pieces of evidence supporting the hierarchical nature of the galaxy formation as predicted by the $\Lambda CDM$ model \citep{1991ApJ...379...52W}. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ on the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of \citet{2024arXiv240519410B} properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

@@

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters by passing closely enough to a segment of a stream and significantly changing the orbits of the closest stars. In this work, \citet{2012ApJ...748...20C} further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided potential observational evidence for such a gap, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters. Interestingly enough the expected mass and size of the perturber causing the gap was inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}.

  Previous studies have predominantly focused on the creation of gaps in stellar streams due to dark matter subhalo passages, with some exceptions. \citet{2016MNRAS.463L..17A} studied the possibility of disc giant molecular clouds gausing gaps in cluster streams. \citet{2017NatAs...1..633P} explored the effects of a galactic bar causing gaps and perturbing other aspects of a stellar stream. To our knowledge, no study has yet been performed investigating the gravitational impact of globular clusters on stellar streams. Since clusters are dense stellar systems, distributed mainly in the Galactic halo, they too should in principle contribute to generating perturbation on stellar streams. If globular clusters can indeed create inhomogeneities and gaps in streams, this factor must be taken into account when using stellar streams to infer the population of dark matter subhaloes.
